---
title: "Alaska 2024 Summer Permafrost Data"
author: "Claire Grenis"
format: html
editor: visual
---

# Alaskan Exploration Data

This is an exploration into the results of the geochemical analysis of samples of thawed permafrost from the Alaskan tundra during the summer months of 2024. Dozens of samples were transported from Alaska to Syracuse University for component analysis including total nitrogen and dissolved organic carbon content, which is what this project will focus on.

### Research Question

The questions that we want to ask and use data visualizations to figure out are:

1.  Is there a relationship between when in the summer this data was recorded and the concentrations of chemical components
2.  Is there a correlation between the concentrations of the two chemical components present in the samples
3.  Are there any hidden relationships that can be uncovered through data visualizations

## Install Packages

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
```

Great! Now we can begin loading in the data sets. Lets use the "glimpse()" function to take a look at what each data set entails.

```{r}
DOC <- read.csv("CGAlaskaDOC.csv")
glimpse(DOC)
```

```{r}
TN <- read.csv("CGAlaskaTN.csv")
glimpse(TN)
```

Each data set, the DOC and the TN, both have three variables.

1.  `Sample ID` Tells us what the individual sample is and is uniquely identifiable
2.  `umol/L C or N` Tells us what the concentration of the respective chemical component being tested for is.
3.  `Date` Tells us when the data was recorded (M/DD/YY format)

## Data Visualizations

Now that we have that all established, we can get into some visualizations and figure out the relationships within the data sets.

First, lets make the variable `umol/L C` and `umol/L N` a little easier to work with.

```{r}
colnames(DOC)[colnames(DOC) == "umol.L.C"] <- "concentration"
colnames(TN)[colnames(TN) == "umol.L.N"] <- "concentration"

```

Now, we have "concentration" as our name to show the amount of carbon or nitrogen in the samples.

Let's also make sure our "Date" is formatted correctly for visualizations. 
```{r}

DOC$Date <- as.Date(DOC$Date, format = "%m/%d/%Y")
TN$Date <- as.Date(TN$Date, format = "%m/%d/%Y")
```

Great! The dates should be alright to work with now. 

First, let's take a look at how to answer the first research question. We need to look at what the relationship between the date that the recording was taken and the concentration of each chemical component is. Two graphs should be produced to demonstrate both carbon and nitrogen.

```{r}
ggplot(DOC, aes(x = Date, y = concentration)) +
  geom_point(color = "tan") +
  geom_smooth(method = "lm", color = "pink", se = FALSE) +
  labs(
    title = "Relationship Between Date and Carbon Concentration",
    x = "Date",
    y = "Carbon Concentration (umol/L)"
  ) +
  theme_minimal()


ggplot(TN, aes(x = Date, y = concentration)) +
  geom_point(color = "tan") +
  geom_smooth(method = "lm", color = "pink", se = FALSE) +
  labs(
    title = "Relationship Between Date and Nitrogen Concentration",
    x = "Date",
    y = "Nitrogen Concentration (umol/L)"
  ) +
  theme_minimal()

```

Now, we want to see what these look like on the same scale to get an idea of what sorts of concentrations we are looking at, relative to each other. 

```{r}

DOC$Type <- "DOC"
TN$Type <- "TN"


combined_data <- bind_rows(DOC, TN)

ggplot(combined_data, aes(x = Date, y = concentration)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Type, ncol = 2) +
  labs(
    title = "Relationship Between Date and Concentration",
    x = "Date",
    y = "Concentration (umol/L)"
  ) +
  theme_minimal()

```
Interesting to see that the concentrations of carbon that we are looking at are so high that they basically make nitrogen obsolete. Let's look at them on the same axes.
```{r}

DOC$Type <- "DOC"
TN$Type <- "TN"
combined_data <- bind_rows(DOC, TN)


ggplot(combined_data, aes(x = Date, y = concentration, color = Type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Date and Concentrations of DOC and TN",
    x = "Date",
    y = "Concentration (umol/L)",
    color = "Chemical Component"
  ) +
  theme_minimal()

```
This gives us the same result, but it is a bit easier to analyze. 

#### 1. It seems that there is a positive correlation between at what point in the warmer months the sample was taken and what concentration in umol/L there is present for both total nitrogen and dissolved organic carbon.

Now we can look to answer the second research question. We want to know if there is a correlation between the concentrations of the two chemical components present in the samples.

```{r}
merged_data <- merge(DOC, TN, by = "Sample.ID", suffixes = c("_DOC", "_TN"))


ggplot(merged_data, aes(x = concentration_DOC, y = concentration_TN)) +
  geom_point(color = "tan") +
  geom_smooth(method = "lm", color = "pink", se = FALSE) +
  labs(
    title = "Correlation Between DOC and TN Concentrations",
    x = "DOC Concentration (umol/L)",
    y = "TN Concentration (umol/L)"
  ) +
  theme_minimal()

```

